---
title: "Homework 3"
author: "Lamiae Taoudi"
date: "2022-11-13"
output: word_document
---

```{r ,message=FALSE, warning=FALSE,setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1: Using data set kyphosis from R package gam

Randomly dividing the data into training (75%) and testing (25%) set

```{r}
library(tidyverse)
library(caret)
data("kyphosis", package = "gam")
head(kyphosis)
dim(kyphosis)

# Split the data into training and test set
set.seed(123)
training.samples <- kyphosis$Age %>%
  createDataPartition(p = 0.75, list = FALSE)
train.data  <- kyphosis[training.samples, ]
test.data <- kyphosis[-training.samples, ]
```

### 1.1 Using function gam to fit a logistic regression model Add nonlinear terms as you see fit
```{r}
library(splines)
library(gam)
gam2 = gam(Kyphosis ~ s(Age, 5) + s(Number, 5) + s(Start,5), family = binomial,data = kyphosis)
par(mfrow = c(1,3))
plot(gam2, se = TRUE, col = "blue")
```
```{r}
summary(gam2)
```


### 1.2 Comparing the testing results of the model built in 1.1 with the results you obtained in HW2 Q2 (2.2)Fitting a logistic regression model: 

```{r}
library(nnet)
# Fit the model
model.logReg <- nnet::multinom(Kyphosis ~., data = train.data)
# Summarize the model
summary(model.logReg)

```
The residual deviance using gam is 35.4366 which is less than the one obtained using the logistic regression model which is 50.3657.


## Question 2
```{r}
n = 100
X  <- runif(n, min = 0, max = 1)
er <- rnorm(n, mean=0, sd=1)

fx <- function(x) sin(12*(x+0.2))/(x+0.2)
Y <- fx(X)+ er

# Fit 2 smoothing splines with df = 5
fit_smooth5 <- smooth.spline(X, Y, df = 5)
fit_smooth_cv <- smooth.spline(X, Y, cv = TRUE)

# Plot the smoothing splines
ggplot() +
  geom_point(aes(x = X, y = Y)) +
  geom_line(aes(x = fit_smooth5$x, y = fit_smooth5$y, 
                color = "5 degrees of freedom"))  +
  geom_line(aes(x = fit_smooth_cv$x, y = fit_smooth_cv$y, 
                color = "CV effective degrees of freedom")) +
  theme(legend.position = 'bottom')+ 
  labs(title = "Smoothing Splines", colour="")
```


```{r}
# Fit 2 smoothing splines with df = 9
fit_smooth9 <- smooth.spline(X, Y, df = 9)

# Plot the smoothing splines
ggplot() +
  geom_point(aes(x = X, y = Y)) +
  geom_line(aes(x = fit_smooth9$x, y = fit_smooth9$y, 
                color = "9 degrees of freedom"))  +
  geom_line(aes(x = fit_smooth_cv$x, y = fit_smooth_cv$y, 
                color = "CV effective degrees of freedom")) +
  theme(legend.position = 'bottom')+ 
  labs(title = "Smoothing Splines", colour="")
```

```{r}
# Fit 2 smoothing splines with df = 15
fit_smooth15 <- smooth.spline(X, Y, df = 15)

# Plot the smoothing splines
ggplot() +
  geom_point(aes(x = X, y = Y)) +
  geom_line(aes(x = fit_smooth15$x, y = fit_smooth15$y, 
                color = "15 degrees of freedom"))  +
  geom_line(aes(x = fit_smooth_cv$x, y = fit_smooth_cv$y, 
                color = "CV effective degrees of freedom")) +
  theme(legend.position = 'bottom')+ 
  labs(title = "Smoothing Splines", colour="")
```

## Question 3: Using data set CanadianWeather from R package "fda"

```{r}
library(fda)                  # load package
data("CanadianWeather")       # load data
attach(CanadianWeather)       # attach data
month <- 1:12
dim(CanadianWeather$monthlyPrecip)
n <- ncol(CanadianWeather$monthlyPrecip)
```

Looking at the monthly precipitation 
```{r}
matplot(month, CanadianWeather$monthlyPrecip,
        type='l', lty=1, col = rainbow(n),
        xlab="months", ylab="precipitation", 
        main="monthly precipitation")
```


### 3.1	Choosing the level of smoothing using generalized cross validation criterion to represent CanadianWeather$monthlyPrecip using a functional object

```{r}
#  -------------------------------------------------------------------
#                 Choose level of smoothing using
#          the generalized cross-validation criterion
#  -------------------------------------------------------------------

# setup basis
mbasis <- create.fourier.basis(rangeval=c(0, 12), nbasis=6)

#  harmonic acceleration operator 
harmaccelLfd <- vec2Lfd(c(0,(2*pi/12)^2,0), c(0, 12))

#  set up range of smoothing parameters in log_10 units

Loglam <- seq(-4,0,0.25)
nglam   <- length(Loglam)

# generate an array to keep track of the results for each iteration
SmoothStats <- array(NA, dim=c(nglam, 3),
      dimnames=list(Loglam, c("log10.lambda", "df", "gcv") ) )
SmoothStats[, 1] <- Loglam

# loop through smoothing parameters

for (ilam in 1:nglam) {
  Smooth <- smooth.basisPar(c(1:12), CanadianWeather$monthlyPrecip, mbasis,
                   Lfdobj=int2Lfd(2), lambda=10^Loglam[ilam])
  SmoothStats[ilam, "df"]  <- Smooth$df
  SmoothStats[ilam, "gcv"] <- sum(Smooth$gcv)
  # note: gcv is a matrix in this case
}

#  display and plot GCV criterion and degrees of freedom
SmoothStats
plot(SmoothStats[, 1], SmoothStats[, 3])
```
The lowest GCV corresponds to log10.lambda= -1 and df =6 
```{r}

plot(Loglam, SmoothStats[, "gcv"], type="b",
     xlab="Log_10 lambda", ylab="GCV Criterion",
     main="Precipitation Smoothing", log="y")

plot(Loglam, SmoothStats[, "df"], type="b",
     xlab="Log_10 lambda", ylab="Degrees of freedom",
     main="Precipitation Smoothing")

```


```{r}
str(CanadianWeather$monthlyPrecip)

precipfd <- smooth.basisPar(c(1:12), CanadianWeather$monthlyPrecip,
       mbasis, Lfdobj=harmaccelLfd, lambda=1e-1)$fd

str(precipfd)

```

### 3.2	Performing functional PCA analysis to CanadianWeather$monthlyPrecip 
```{r}
#  --------------------------------------------------------------
#            Principal components analysis
#  --------------------------------------------------------------

#  do the PCA with varimax rotation

# Smooth with lambda as determined above
fdPar  <- fdPar(mbasis, harmaccelLfd, lambda=1e-1)
pca.fd <- pca.fd(precipfd, nharm=4, fdPar)

pca.fd <- varmx.pca.fd(pca.fd)

#  plot harmonics using cycle plots

#par(mfrow=c(1,1), mar=c(3,4,2,1), pty="s")
op <- par(mfrow=c(2,2))
plot.pca.fd(pca.fd, cycle=TRUE)
par(op)
```


